---
title: "20231214_RoundRobin_reproducibility_template"
author: "Haley"
date: "2023-12-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code determines the inter-lab variability of features detected using ClusterFinder across labs

Comparing:
- RSD of features in normalized, SC, and raw datasets
- Pearson correlations of shared features
- linear model association of shared features
- linear model associations of shared features, controlled for method variables

Input file structure:

- parent folder path structure:
  "/Users/chatelaineha/Documents/Beecher_Collaboration/CF_projects"

- IROA results file names:
  "path/Results/[* Norm/SC/Raw *.xlsx]"
  
```{r}
source("./code/plotting_functions.R")
source("./code/reproducibility_functions.R")
source("./code/summary_functions.R")

library(devtools)
#install_github("jokergoo/InteractiveComplexHeatmap")
#library(InteractiveComplexHeatmap)
```

# Load data

```{r}
load("./data/reproducibility_input/duplicated_filtered_data.RData")
```

# Determine ClassyFire super class of metabolites using RaMP-DB 2.0 (for use in heatmap annotations)

```{r}
# Load RaMP-DB
if (!requireNamespace("ncats/RAMP-DB", quietly = TRUE)) {
  library(devtools)
  #install_github("ncats/RAMP-DB")
  pak::pak("ncats/RaMP-DB")

}

invisible(lapply(packages, library, character.only = TRUE))
library(RaMP)
library(DT) 
library(dplyr)
library(magrittr)

rampDB <- RaMP()

comp.class <- map.classes(comp.std)
```

# Find metabolites that overlap across all labs:

```{r}
retain.name.results.overlap <- overlap.metabs(comp.std, "retain.name")
print(paste(length(retain.name.results.overlap), " metabolites overlap by retained name across all 8 labs"))

name.results.overlap <- overlap.metabs(comp.std, "Name")
print(paste(length(name.results.overlap), " metabolites overlap by name across all 8 labs"))

form.results.overlap <- overlap.metabs(comp.std, "MF")
print(paste(length(form.results.overlap), " metabolites overlap by formula across all 8 labs"))

```

# Qualitative Overlaps

## Make UpSet plot by metabolite name
### subset by match type 2 and 3a**

```{r, results='asis'}
# Create data for upset
upset.metabs <- upset.prep(comp.class, "retain.name")

## set aside an object with class names as characters for bar plot later
upset.metabs.classnames <- upset.metabs

# Prep colors

mycolors <- graf_palettes$kelly[1:length(unique(upset.metabs$class))]
names(mycolors) <- unique(upset.metabs$class)

# Convert class to numeric factors

upset.metabs$class <- as.numeric(as.factor(upset.metabs$class))

# Set colors to match class factors for upset

class_factor <- as.factor(names(mycolors))

# create query for class colors

mylist<-list()
i<-0
while(length(class_factor)>0){
  i <- i+1
  mylist[[i]] <- list(query = elements, 
                      params = list("class",class_factor),
                      color = mycolors[i],
                      active = T)
  class_factor <- class_factor[-1]
}

# plot

pdf("./outputs/final_metabs_upset_retainname.pdf",width = 32,height = 16,fonts = "serif")
class_upset <- upset(upset.metabs, 
                    queries = mylist,
                    nsets = length(comp.class),
                    order.by = "degree", 
                    text.scale = 3, 
                    point.size = 5,
                    nintersects = NA,
                    keep.order = T)

class_upset

# find how many metabolites are in each "bin" of lab presence

upset.metabs.classnames$bin <- rowSums(subset(upset.metabs.classnames, select = which(grepl("Lab", colnames(upset.metabs.classnames)))))

table(upset.metabs.classnames$bin)
```

## Make bar plot of total classes across all labs

```{r}
# Prep data
metab.bar <- upset.metabs.classnames %>% group_by(class) %>% mutate(count = n())

metab.bar$class <- as.factor(metab.bar$class)

# Prep colors

mycolors <- graf_palettes$kelly[1:length(unique(metab.bar$class))]
names(mycolors) <- unique(metab.bar$class)

# Plot

class_bar <- ggplot(data = metab.bar, aes(x = reorder(class, count), fill = class)) +
              geom_bar() +
              scale_fill_manual(values = mycolors) +
              coord_flip() +
              theme_classic() +
              theme(legend.position = "none")

class_bar

ggsave("./outputs/total_classes.png", width = 6, height = 4)
```

# Metabolite Correlations

## WITH Lab 2

### Calculate pairwise correlations of RATIO data for metabolites that overlap (by name) across all labs using any match type WITH lab 2

```{r}
# Subset metabs to use
ratio.30.subset <- lapply(ratio.std, function(x) x[,which(colnames(x) %in% retain.name.results.overlap)])

# Shorten long metabolite names
short.names <- function(lab, nchar, ratio.30.subset, comp.std){
  # lab = character vector of lab names
  # nchar = length of metabolite name to shorten (number of characters)
  # ratio.30.subset = list of subset ratio data from all labs
  # comp.std = list of metabolite meta data from all labs
  
  longs <- colnames(ratio.30.subset[[lab]])[nchar(colnames(ratio.30.subset[[lab]])) > 20]

match_idx <- match(longs, comp.std[[lab]]$retain.name)
colnames(ratio.30.subset[[lab]])[which(colnames(ratio.30.subset[[lab]]) %in% longs)] <- comp.std[[lab]]$MF[match_idx]
 
return(ratio.30.subset[[lab]])
 
}

simple.30.subset <- lapply(names(ratio.30.subset), function(x) short.names(x, 20, ratio.30.subset, comp.std))

names(simple.30.subset) <- names(ratio.30.subset)

metabs.30 <- colnames(simple.30.subset$Lab1)


# log transform
simple.30.subset.log <- lapply(simple.30.subset, log2)

# correlation
simple.30.subset.cor <- lab.cor(simple.30.subset.log, "pearson")
simple.30.subset.cor <- simple.30.subset.cor[order(row.names(simple.30.subset.cor)),]

# heatmap
png(file = "./outputs/correlation_heatmap_30_alltypes_simple_withLab2.png", width = 8, height = 12, units = "in", res = 300)
heat.cor(simple.30.subset.cor, show.colnames = T, show.rownames = T, show.negatives = T, fontsize = 10)

# summary
simple.30.subset.mean <- mean(unlist(simple.30.subset.cor), na.rm = T)
n <- length(simple.30.subset.cor)
std.dev.simple.30.subset <- sd(unlist(simple.30.subset.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) simple data have a mean of", round(simple.30.subset.mean, 2), "and standard deviation of", round(std.dev.simple.30.subset, 2)))
```

### Calculate pairwise correlations of RATIO data for metabolites that have standard data in 3+ labs using any match type WITH lab 2

```{r}
# Subset metabs to use
ratio.subset.stds <- mapply(function(x, y) match.subset(x, y, "2", is.data = T), ratio.std, comp.std)
std.metabs <- colnames(ratio.subset.stds$Lab1)
ratio.std.subset <- lapply(ratio.std, function(x) x[,which(colnames(x) %in% std.metabs)])

# log transform
ratio.std.subset.log <- lapply(ratio.std.subset, log2)

# correlation
ratio.std.subset.cor <- lab.cor(ratio.std.subset.log, "pearson")
ratio.std.subset.cor <- ratio.std.subset.cor[order(row.names(ratio.std.subset.cor)),]

# heatmap
png(file = "./outputs/correlation_heatmap_stds_alltypes_ratio_withLab2.png", width = 3, height = 8, units = "in", res = 300)
heat.cor(ratio.std.subset.cor, show.colnames = T, show.rownames = F, show.negatives = F)

# summary
ratio.std.subset.mean <- mean(unlist(ratio.std.subset.cor), na.rm = T)
n <- length(ratio.std.subset.cor)
std.dev.ratio.std.subset <- sd(unlist(ratio.std.subset.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) ratio data have a mean of", round(ratio.std.subset.mean, 2), "and standard deviation of", round(std.dev.ratio.std.subset, 2)))
```

### Calculate pairwise correlations of NORM data for metabolites that have standard data in 3+ labs using any match type WITH lab 2

```{r}
# Subset metabs to use
norm.subset.stds <- mapply(function(x, y) match.subset(x, y, "2", is.data = T), norm.std, comp.std)
std.metabs <- colnames(norm.subset.stds$Lab1)
norm.std.subset <- lapply(norm.std, function(x) x[,which(colnames(x) %in% std.metabs)])

# log transform
norm.std.subset.log <- lapply(norm.std.subset, log2)

# correlation
norm.std.subset.cor <- lab.cor(norm.std.subset.log, "pearson")
norm.std.subset.cor <- norm.std.subset.cor[order(row.names(norm.std.subset.cor)),]

# heatmap
png(file = "./outputs/correlation_heatmap_stds_alltypes_norm_withLab2.png", width = 4, height = 8, units = "in", res = 300)
heat.cor(norm.std.subset.cor, show.colnames = T, show.rownames = T, show.negatives = F)

# summary
norm.std.subset.mean <- mean(unlist(norm.std.subset.cor), na.rm = T)
n <- length(norm.std.subset.cor)
std.dev.norm.std.subset <- sd(unlist(norm.std.subset.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) norm data have a mean of", round(norm.std.subset.mean, 2), "and standard deviation of", round(std.dev.norm.std.subset, 2)))
```

### Calculate pairwise correlations of MSTUS data for metabolites that have standard data in 3+ labs using any match type WITH lab 2

```{r}
# Subset metabs to use
mstus.subset.stds <- mapply(function(x, y) match.subset(x, y, "2", is.data = T), mstus.std, comp.std)
std.metabs <- colnames(mstus.subset.stds$Lab1)
mstus.std.subset <- lapply(mstus.std, function(x) x[,which(colnames(x) %in% std.metabs)])

# log transform
mstus.std.subset.log <- lapply(mstus.std.subset, log2)

# correlation
mstus.std.subset.cor <- lab.cor(mstus.std.subset.log, "pearson")
mstus.std.subset.cor <- mstus.std.subset.cor[order(row.names(mstus.std.subset.cor)),]

# heatmap
png(file = "./outputs/correlation_heatmap_stds_alltypes_mstus_withLab2.png", width = 3, height = 8, units = "in", res = 300)
heat.cor(mstus.std.subset.cor, show.colnames = T, show.rownames = F, show.negatives = F)

# summary
mstus.std.subset.mean <- mean(unlist(mstus.std.subset.cor), na.rm = T)
n <- length(mstus.std.subset.cor)
std.dev.mstus.std.subset <- sd(unlist(mstus.std.subset.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) mstus data have a mean of", round(mstus.std.subset.mean, 2), "and standard deviation of", round(std.dev.mstus.std.subset, 2)))
```

### Calculate pairwise correlations of RAW data for metabolites that have standard data in 3+ labs using any match type WITH lab 2

```{r}
# Subset metabs to use
raw.subset.stds <- mapply(function(x, y) match.subset(x, y, "2", is.data = T), raw.std, comp.std)
std.metabs <- colnames(raw.subset.stds$Lab1)
raw.std.subset <- lapply(raw.std, function(x) x[,which(colnames(x) %in% std.metabs)])

# log transform
raw.std.subset.log <- lapply(raw.std.subset, log2)

# correlation
raw.std.subset.cor <- lab.cor(raw.std.subset.log, "pearson")
raw.std.subset.cor <- raw.std.subset.cor[order(row.names(raw.std.subset.cor)),]

# heatmap
png(file = "./outputs/correlation_heatmap_stds_alltypes_raw_withLab2.png", width = 3, height = 8, units = "in", res = 300)
heat.cor(raw.std.subset.cor, show.colnames = T, show.rownames = F, show.negatives = F)

# summary
raw.std.subset.mean <- mean(unlist(raw.std.subset.cor), na.rm = T)
n <- length(raw.std.subset.cor)
std.dev.raw.std.subset <- sd(unlist(raw.std.subset.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) raw data have a mean of", round(raw.std.subset.mean, 2), "and standard deviation of", round(std.dev.raw.std.subset, 2)))
```

## WITHOUT Lab 2

### Calculate pairwise correlations of RATIO data for metabolites that have standard data in 3+ labs using any match type WITHOUT lab 2

```{r}
# Subset metabs to use
std.metabs <- colnames(ratio.subset.stds$Lab1)
ratio.std.subset <- lapply(ratio.std, function(x) x[,which(colnames(x) %in% std.metabs)])

# Remove Lab 2
ratio.std.subset.no2 <- ratio.std.subset[-which(names(ratio.std.subset) == "Lab2")]

# log transform
ratio.std.subset.no2.log <- lapply(ratio.std.subset.no2, log2)

# correlation
ratio.std.subset.no2.cor <- lab.cor(ratio.std.subset.no2.log, "pearson")

# heatmap
png(file = "./outputs/correlation_heatmap_stds_alltypes_ratio.png", width = 3, height = 8, units = "in", res = 300)
heat.cor(ratio.std.subset.no2.cor, show.colnames = T, show.rownames = F, show.negatives = F)

ratio.std.heat.no2 <- heat.cor(ratio.std.subset.no2.cor, show.colnames = T, show.rownames = F, legend = F, show.negatives = F)

# summary
ratio.std.subset.mean <- mean(unlist(ratio.std.subset.no2.cor), na.rm = T)
n <- length(ratio.std.subset.no2.cor)
std.dev.ratio.std.subset <- sd(unlist(ratio.std.subset.no2.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) ratio data have a mean of", round(ratio.std.subset.mean, 2), "and standard deviation of", round(std.dev.ratio.std.subset, 2)))
```

### Calculate pairwise correlations of NORM data for metabolites that have standard data in 3+ labs using any match type WITHOUT lab 2

```{r}
# Subset metabs to use
norm.std.subset <- lapply(norm.std, function(x) x[,which(colnames(x) %in% std.metabs)])

# Remove Lab 2
norm.std.subset.no2 <- norm.std.subset[-which(names(norm.std.subset) == "Lab2")]

# log transform
norm.std.subset.no2.log <- lapply(norm.std.subset.no2, log2)

# correlation
norm.std.subset.no2.cor <- lab.cor(norm.std.subset.no2.log, "pearson")

# heatmap
png(file = "./outputs/correlation_heatmap_stds_alltypes_norm.png", width = 4, height = 8, units = "in", res = 300)
heat.cor(norm.std.subset.no2.cor, show.rownames = T, show.colnames = T, show.negatives = F)

norm.std.heat.no2 <- heat.cor(norm.std.subset.no2.cor, show.colnames = T, show.negatives = F)

# summary
norm.std.subset.mean <- mean(unlist(norm.std.subset.no2.cor), na.rm = T)
n <- length(norm.std.subset.no2.cor)
std.dev.norm.std.subset <- sd(unlist(norm.std.subset.no2.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) norm data have a mean of", round(norm.std.subset.mean, 2), "and standard deviation of", round(std.dev.norm.std.subset, 2)))
```

### Calculate pairwise correlations of MSTUS data for metabolites that have standard data in 3+ labs using any match type WITHOUT lab 2

```{r}
# Subset metabs to use
mstus.std.subset <- lapply(mstus.std, function(x) x[,which(colnames(x) %in% std.metabs)])

# Remove Lab 2
mstus.std.subset.no2 <- mstus.std.subset[-which(names(mstus.std.subset) == "Lab2")]

# log transform
mstus.std.subset.no2.log <- lapply(mstus.std.subset.no2, log2)

# correlation
mstus.std.subset.no2.cor <- lab.cor(mstus.std.subset.no2.log, "pearson")

# heatmap
png(file = "./outputs/correlation_heatmap_stds_alltypes_mstus.png", width = 2.5, height = 8, units = "in", res = 300)
heat.cor(mstus.std.subset.no2.cor, show.colnames = T, show.rownames = F, show.negatives = F)

mstus.std.heat.no2 <- heat.cor(mstus.std.subset.no2.cor, show.colnames = T, show.rownames = F, legend = F, show.negatives = F)

# summary
mstus.std.subset.mean <- mean(unlist(mstus.std.subset.no2.cor), na.rm = T)
n <- length(mstus.std.subset.no2.cor)
std.dev.mstus.std.subset <- sd(unlist(mstus.std.subset.no2.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) mstus data have a mean of", round(mstus.std.subset.mean, 2), "and standard deviation of", round(std.dev.mstus.std.subset, 2)))
```

### Calculate pairwise correlations of SC data for metabolites that have standard data in 3+ labs using any match type WITHOUT lab 2

```{r}
# Subset metabs to use
sc.std.subset <- lapply(sc.std, function(x) x[,which(colnames(x) %in% std.metabs)])

# Remove Lab 2
sc.std.subset.no2 <- sc.std.subset[-which(names(sc.std.subset) == "Lab2")]

# log transform
sc.std.subset.no2.log <- lapply(sc.std.subset.no2, log2)

# correlation
sc.std.subset.no2.cor <- lab.cor(sc.std.subset.no2.log, "pearson")

# heatmap
png(file = "./outputs/correlation_heatmap_stds_alltypes_sc.png", width = 3, height = 8, units = "in", res = 300)
heat.cor(sc.std.subset.no2.cor, show.rownames = F, show.colnames = T, show.negatives = F)

sc.std.heat.no2 <- heat.cor(sc.std.subset.no2.cor, show.colnames = T, show.rownames = F, legend = F, show.negatives = F)

# summary
sc.std.subset.mean <- mean(unlist(sc.std.subset.no2.cor), na.rm = T)
n <- length(sc.std.subset.no2.cor)
std.dev.sc.std.subset <- sd(unlist(sc.std.subset.no2.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) sc data have a mean of", round(sc.std.subset.mean, 2), "and standard deviation of", round(std.dev.sc.std.subset, 2)))
```

### Calculate pairwise correlations of RAW data for metabolites that have standard data in 3+ labs using any match type WITHOUT lab 2

```{r}
# Subset metabs to use
raw.std.subset <- lapply(raw.std, function(x) x[,which(colnames(x) %in% std.metabs)])

# Remove Lab 2
raw.std.subset.no2 <- raw.std.subset[-which(names(raw.std.subset) == "Lab2")]

# log transform
raw.std.subset.no2.log <- lapply(raw.std.subset.no2, log2)

# correlation
raw.std.subset.no2.cor <- lab.cor(raw.std.subset.no2.log, "pearson")

# heatmap
png(file = "./outputs/correlation_heatmap_stds_alltypes_raw.png", width = 2.5, height = 8, units = "in", res = 300)
heat.cor(raw.std.subset.no2.cor, show.rownames = F, show.colnames = T, show.negatives = F)

raw.std.heat.no2 <- heat.cor(raw.std.subset.no2.cor, show.colnames = T, show.rownames = F, legend = F, show.negatives = F)

# summary
raw.std.subset.mean <- mean(unlist(raw.std.subset.no2.cor), na.rm = T)
n <- length(raw.std.subset.no2.cor)
std.dev.raw.std.subset <- sd(unlist(raw.std.subset.no2.cor), na.rm = T)

print(paste("The correlations in the standard-subset (all match types) raw data have a mean of", round(raw.std.subset.mean, 2), "and standard deviation of", round(std.dev.raw.std.subset, 2)))
```

## All metabolites

### Calculate pairwise correlations of abundance data for shared metabolites (by name) between labs - Pearson

```{r, message=FALSE}
# log2 transform data
norm.log <- lapply(norm.std, log2)
ratio.log <- lapply(ratio.std, log2)
mstus.log <- lapply(mstus.std, log2)

# pearson - no dupes dataset
norm.pearson <- lab.cor(norm.log, "pearson")
ratio.pearson <- lab.cor(ratio.log, "pearson")
mstus.pearson <- lab.cor(mstus.log, "pearson")

norm.cor.heat <- heat.cor(norm.pearson)

#norm.cor.heat.class <- heat.cor(norm.pearson, comp.class, order.by.class = T)
#norm.cor.heat.class
#htShiny()
#sc.cor.heat <- heat.cor(sc.pearson)
#raw.cor.heat <- heat.cor(raw.pearson)

violin.norm <- violin.cor(norm.pearson, "normalized")
violin.norm
#ggsave("./outputs/violin_norm_feature_correlations.png", width = 6, height = 4, dpi = 600)

violin.ratio <- violin.cor(ratio.pearson, "ratio")
violin.ratio
#ggsave("./outputs/violin_ratio_feature_correlations.png", width = 6, height = 4, dpi = 600)

violin.mstus <- violin.cor(mstus.pearson, "mstus")
violin.mstus
#ggsave("./outputs/violin_ratio_feature_correlations.png", width = 6, height = 4, dpi = 600)

#arranged.nodup.violin <- ggarrange(violin.raw, violin.sc, violin.norm, ncol = 1, nrow = 3)
#arranged.nodup.violin

# heatmaps
png(file = "./outputs/correlation_heatmap_allmetabs_norm.png", width = 8, height = 6, units = "in", res = 300)
heat.cor(norm.pearson)

png(file = "./outputs/correlation_heatmap_allmetabs_ratio.png", width = 8, height = 6, units = "in", res = 300)
heat.cor(ratio.pearson)

png(file = "./outputs/correlation_heatmap_allmetabs_mstus.png", width = 8, height = 6, units = "in", res = 300)
heat.cor(mstus.pearson)

# NORMALIZED

norm.pearson.mean <- mean(unlist(norm.pearson), na.rm = T)
n <- length(norm.pearson)
std.dev.norm <- sd(unlist(norm.pearson), na.rm = T)

margin.error <- qt(0.975, df = n-1) * std.dev.norm/sqrt(n)

lower.ci <- norm.pearson.mean - margin.error
upper.ci <- norm.pearson.mean + margin.error

print(paste("The correlations in the normalized data have a mean of", round(norm.pearson.mean, 2), "and standard deviation of", round(std.dev.norm, 2)))

print(paste("The correlations in the normalized data have a trimmed (10%) mean of", round(mean(unlist(norm.pearson), na.rm = T, trim = 0.1), 2)))

# RATIO

ratio.pearson.mean <- mean(unlist(ratio.pearson), na.rm = T)
n <- length(ratio.pearson)
std.dev.ratio <- sd(unlist(ratio.pearson), na.rm = T)

margin.error <- qt(0.975, df = n-1) * std.dev.ratio/sqrt(n)

lower.ci <- ratio.pearson.mean - margin.error
upper.ci <- ratio.pearson.mean + margin.error

print(paste("The correlations in the ratio data have a mean of", round(ratio.pearson.mean, 2), "and standard deviation of", round(std.dev.ratio, 2)))

print(paste("The correlations in the ratio data have a trimmed (10%) mean of", round(mean(unlist(ratio.pearson), na.rm = T, trim = 0.1), 2)))

# MSTUS

mstus.pearson.mean <- mean(unlist(mstus.pearson), na.rm = T)
n <- length(mstus.pearson)
std.dev.mstus <- sd(unlist(mstus.pearson), na.rm = T)

margin.error <- qt(0.975, df = n-1) * std.dev.mstus/sqrt(n)

lower.ci <- mstus.pearson.mean - margin.error
upper.ci <- mstus.pearson.mean + margin.error

print(paste("The correlations in the mstus data have a mean of", round(mstus.pearson.mean, 2), "and standard deviation of", round(std.dev.mstus, 2)))

print(paste("The correlations in the mstus data have a trimmed (10%) mean of", round(mean(unlist(mstus.pearson), na.rm = T, trim = 0.1), 2)))

```

### Check distribution of negative values by lab pair (use full comparisons)

```{r}
#mstus.lab.negs <- apply(mstus.pearson, 2, function (x) length(which(x < 0)))
#plot(mstus.lab.negs ~ names(mstus.lab.negs))

```


# PCAs with NON-duplicate-filtered data

```{r}
load("./data/duplicate_input/imputed_data.RData")
```

## color by group and shape by concentration

```{r}
# set group colors with gradients within groups
group.colors <- c("red","orange", "brown","purple","gold", "darkgreen","blue")
names(group.colors) <- levels(meta.data$group)

# normalized data - group and concentration
norm.pca.list <- list()
for(i in 1:length(norm.impute)){
  
  pca.i <- make.pca(norm.impute[[i]], meta.data, class = "group", group.shape.name = "concentration", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(norm.impute)[[i]], " SC Normalized"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      norm.pca.list[[i]] <- pca.i
    }else{
      norm.pca.list[[length(norm.pca.list) + 1]] <- pca.i
    }
}

norm.arranged <- ggarrange(plotlist = norm.pca.list, ncol = 8, nrow = 1, common.legend = T) 
ggsave("./outputs/pca_groupconc_shape_normalized_WITHdupes.png", norm.arranged, width = 50, height = 10, dpi = 600, limitsize = F)
norm.arranged

# normalized data - group only
norm.pca.list <- list()
for(i in 1:length(norm.impute)){
  
  pca.i <- make.pca(norm.impute[[i]], meta.data, class = "group", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(norm.impute)[[i]], " SC Normalized"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      norm.pca.list[[i]] <- pca.i
    }else{
      norm.pca.list[[length(norm.pca.list) + 1]] <- pca.i
    }
}

norm.arranged <- ggarrange(plotlist = norm.pca.list, ncol = 4, nrow = 2, common.legend = T) 
ggsave("./outputs/pca_group_normalized_WITHdupes.png", norm.arranged, width = 30, height = 15, dpi = 600)

# sc data - group and concentration
sc.pca.list <- list()
for(i in 1:length(sc.impute)){
  
  pca.i <- make.pca(sc.impute[[i]], meta.data, class = "group", group.shape.name = "concentration", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(sc.impute)[[i]], " Suppression Corrected"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      sc.pca.list[[i]] <- pca.i
    }else{
      sc.pca.list[[length(sc.pca.list) + 1]] <- pca.i
    }
}

sc.arranged <- ggarrange(plotlist = sc.pca.list, ncol = 8, nrow = 1, common.legend = T) 
ggsave("./outputs/pca_groupconc_shape_sc_WITHdupes.png", sc.arranged, width = 50, height = 10, dpi = 600, limitsize = F)
sc.arranged

# sc data - group only
sc.pca.list <- list()
for(i in 1:length(sc.impute)){
  
  pca.i <- make.pca(sc.impute[[i]], meta.data, class = "group", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(sc.impute)[[i]], " suppression corrected"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      sc.pca.list[[i]] <- pca.i
    }else{
      sc.pca.list[[length(sc.pca.list) + 1]] <- pca.i
    }
}

sc.arranged <- ggarrange(plotlist = sc.pca.list, ncol = 4, nrow = 2, common.legend = T) 
ggsave("./outputs/pca_group_sc_WITHdupes.png", sc.arranged, width = 30, height = 15, dpi = 600)

# raw data - group and concentration
raw.pca.list <- list()
for(i in 1:length(raw.impute)){
  
  pca.i <- make.pca(raw.impute[[i]], meta.data, class = "group", group.shape.name = "concentration", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(raw.impute)[[i]], " Clean"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      raw.pca.list[[i]] <- pca.i
    }else{
      raw.pca.list[[length(raw.pca.list) + 1]] <- pca.i
    }
}

raw.arranged <- ggarrange(plotlist = raw.pca.list, ncol = 8, nrow = 1, common.legend = T) 
ggsave("./outputs/pca_groupconc_shape_raw_WITHdupes.png", raw.arranged, width = 50, height = 10, dpi = 600, limitsize = F)
raw.arranged

# raw data - group only
raw.pca.list <- list()
for(i in 1:length(raw.impute)){
  
  pca.i <- make.pca(raw.impute[[i]], meta.data, class = "group", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(raw.impute)[[i]], " rawalized"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      raw.pca.list[[i]] <- pca.i
    }else{
      raw.pca.list[[length(raw.pca.list) + 1]] <- pca.i
    }
}

raw.arranged <- ggarrange(plotlist = raw.pca.list, ncol = 4, nrow = 2, common.legend = T) 
ggsave("./outputs/pca_group_raw_WITHdupes.png", raw.arranged, width = 30, height = 15, dpi = 600)

# mstus data - group and concentration
mstus.pca.list <- list()
for(i in 1:length(mstus.impute)){
  
  pca.i <- make.pca(mstus.impute[[i]], meta.data, class = "group", group.shape.name = "concentration", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(mstus.impute)[[i]], " Clean Normalized"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      mstus.pca.list[[i]] <- pca.i
    }else{
      mstus.pca.list[[length(mstus.pca.list) + 1]] <- pca.i
    }
}

mstus.arranged <- ggarrange(plotlist = mstus.pca.list, ncol = 8, nrow = 1, common.legend = T) 
ggsave("./outputs/pca_groupconc_shape_mstus_WITHdupes.png", mstus.arranged, width = 50, height = 10, dpi = 600, limitsize = F)
mstus.arranged

# mstus data - group only
mstus.pca.list <- list()
for(i in 1:length(mstus.impute)){
  
  pca.i <- make.pca(mstus.impute[[i]], meta.data, class = "group", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(mstus.impute)[[i]], " mstusalized"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      mstus.pca.list[[i]] <- pca.i
    }else{
      mstus.pca.list[[length(mstus.pca.list) + 1]] <- pca.i
    }
}

mstus.arranged <- ggarrange(plotlist = mstus.pca.list, ncol = 4, nrow = 2, common.legend = T) 
ggsave("./outputs/pca_group_mstus_WITHdupes.png", mstus.arranged, width = 30, height = 15, dpi = 600)

# ratio data - group and concentration
ratio.pca.list <- list()
for(i in 1:length(ratio.impute)){
  
  pca.i <- make.pca(ratio.impute[[i]], meta.data, class = "group", group.shape.name = "concentration", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(ratio.impute)[[i]], " Ratio"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      ratio.pca.list[[i]] <- pca.i
    }else{
      ratio.pca.list[[length(ratio.pca.list) + 1]] <- pca.i
    }
}

ratio.arranged <- ggarrange(plotlist = ratio.pca.list, ncol = 8, nrow = 1, common.legend = T) 
ggsave("./outputs/pca_groupconc_shape_ratio_WITHdupes.png", ratio.arranged, width = 50, height = 10, dpi = 600, limitsize = F)
ratio.arranged

# ratio data - group only
ratio.pca.list <- list()
for(i in 1:length(ratio.impute)){
  
  pca.i <- make.pca(ratio.impute[[i]], meta.data, class = "group", do.log2 = T, centered = T, scaled = T, data.type = paste0(names(ratio.impute)[[i]], " ratioalized"), point.size = 8, text.size = 25, color.vector = group.colors, z.axis = F)
  
      if(i == 1){
      ratio.pca.list[[i]] <- pca.i
    }else{
      ratio.pca.list[[length(ratio.pca.list) + 1]] <- pca.i
    }
}

ratio.arranged <- ggarrange(plotlist = ratio.pca.list, ncol = 4, nrow = 2, common.legend = T) 
ggsave("./outputs/pca_group_ratio_WITHdupes.png", ratio.arranged, width = 30, height = 15, dpi = 600)
```

## Find important features in PCAs of groups

```{r}
# normalized
norm.group.load <- list(length(norm.impute))

for(i in 1:length(norm.impute)){
  
  norm.group.load[[i]] <- feat.pca(norm.impute[[i]], meta.data, "group", do.log2 = T, centered = T, scaled = T, z.axis = T)
  
  row.names(norm.group.load[[i]]) <- row.names(comp.found.na.filter[[i]])
  
}

names(norm.group.load) <- names(norm.impute)

# save for biomarker analysis
save(norm.group.load, comp.found.na.filter, file = "./data/biomarker_input/PCA_features_WITHduplicates.RData")
```

# Linear models: confirm associations with group and concentration across data types - metabolites

## Function to write linear models and calculate sum-of-squares for group and conc contributions

```{r}
lm.group.conc <- function(data.list, meta.data, summarize = TRUE, pvals = FALSE){
  # data.list = metabolite abundance data for all labs
  # meta.data = sample meta data
  # summarize = logical indicating whether to output simplified summary (TRUE) or full LM contribution to variance results (FALSE)
  # pvals = logical indicating whether to output pvalues associated with each coefficient
  
  # function for one lab
    lm.one.lab <- function(data, meta.data, summarize = summarize){
    ## run model and extract results
    model <- apply(data, 2, function(x) lm(x ~ meta.data$group + meta.data$concentration))
    
    anova.res <- lapply(model, function(x) Anova(x, type = "II"))
    
    results <- lapply(anova.res, function(x) data.frame(Term = rownames(x),
                                                        SumSq = x$`Sum Sq`,
                                                        PctContribution = round(100 * x$`Sum Sq` / sum(x$`Sum Sq`), 1),
                                                        PValue = x$`Pr(>F)`
                                                        ))
    ## pull results into clean table

    # Combine into one long data frame, tagging each row with its metabolite
    results_long <- do.call(rbind, lapply(names(results), function(m) {
      df <- results[[m]]
      df$Metabolite <- m
      df
    }))

    # Clean up term names (they'll be like "meta.data$group", "meta.data$concentration", "Residuals")
    results_long$Term <- gsub("meta\\.data\\$", "", results_long$Term)

    if(summarize == TRUE){
      
    # export cleaned summary values  
      
    # Summarize: mean and sd of PctContribution per term, across all metabolites
    summary_stats <- results_long %>%
      group_by(Term) %>%
      summarise(Mean = mean(PctContribution, na.rm = TRUE),
                SD = sd(PctContribution, na.rm = TRUE))
    
    # Reshape into the wide format you want: one row, columns = group/concentration/Residuals as "mean ± sd"
    final_result <- data.frame(
      group = sprintf("%.1f ± %.1f", 
                       summary_stats$Mean[summary_stats$Term == "group"],
                       summary_stats$SD[summary_stats$Term == "group"]),
      concentration = sprintf("%.1f ± %.1f", 
                               summary_stats$Mean[summary_stats$Term == "concentration"],
                               summary_stats$SD[summary_stats$Term == "concentration"]),
      Residuals = sprintf("%.1f ± %.1f", 
                           summary_stats$Mean[summary_stats$Term == "Residuals"],
                           summary_stats$SD[summary_stats$Term == "Residuals"])
    )
      
          if(pvals == TRUE){
        pval_summary <- results_long %>%
          filter(Term %in% c("group", "concentration")) %>%
          group_by(Term) %>%
          summarise(MeanP = mean(PValue, na.rm = TRUE),
                    MedianP = median(PValue, na.rm = TRUE),
                    PctSig = round(100 * mean(PValue < 0.05, na.rm = TRUE), 1),
                    .groups = "drop")

        final_result$group_pval <- sprintf(
          "mean=%.3g, median=%.3g, %%sig=%.1f",
          pval_summary$MeanP[pval_summary$Term == "group"],
          pval_summary$MedianP[pval_summary$Term == "group"],
          pval_summary$PctSig[pval_summary$Term == "group"]
        )
        final_result$concentration_pval <- sprintf(
          "mean=%.3g, median=%.3g, %%sig=%.1f",
          pval_summary$MeanP[pval_summary$Term == "concentration"],
          pval_summary$MedianP[pval_summary$Term == "concentration"],
          pval_summary$PctSig[pval_summary$Term == "concentration"]
        )
      }
    
    } else {
    
      if(pvals == TRUE){
        final_result <- results_long %>%
          select(Metabolite, Term, PctContribution, PValue) %>%
          pivot_wider(
            names_from = Term,
            values_from = c(PctContribution, PValue)
          )
        # columns will be named e.g. PctContribution_group, PValue_group, etc.
      } else {
        final_result <- results_long %>%
          select(Metabolite, Term, PctContribution) %>%
          pivot_wider(
            names_from = Term,
            values_from = PctContribution
          )
      }

    }
     return(final_result) 
    }
    
    # apply to all labs
    
  result_all <- lapply(data.list, function(x) lm.one.lab(x, meta.data, summarize))
  
  for(lab in 1:length(result_all)){
    
    result_all[[lab]]$lab <- names(result_all)[[lab]]
    
  }
  
  fin_result_all <- do.call(rbind, result_all)

  return(fin_result_all)
  
}
```

### LM: Metabolite ~ group + concentration

#### Apply LM function to data types and summarize results

```{r}
raw.lm <- lm.group.conc(lapply(raw.impute, log2), meta.data, summarize = T, pvals = T)
sc.lm <- lm.group.conc(lapply(sc.impute, log2), meta.data, summarize = T, pvals = T)
mstus.lm <- lm.group.conc(lapply(mstus.impute, log2), meta.data, summarize = T, pvals = T)
norm.lm <- lm.group.conc(lapply(norm.impute, log2), meta.data, summarize = T, pvals = T)
```

#### Plot relative contributions per data type (stacked bar plot)

```{r}
extract_mean <- function(x) as.numeric(sub(" ±.*", "", x))

extract_sd <- function(x) as.numeric(sub("^.*± ", "", x))

summarize_table <- function(df, dtype_name) {
  data.frame(
    DataType = dtype_name,
    Term = c("group", "concentration", "residuals"),
    Mean = c(mean(extract_mean(df$group)),
             mean(extract_mean(df$concentration)),
             mean(extract_mean(df$Residuals))),
    SD = c(round(mean(extract_sd(df$group)), 0),
             round(mean(extract_sd(df$concentration)), 0),
             round(mean(extract_sd(df$Residuals)), 0))    
  )
}

# df1, df2, df3, df4 = your four Lab1-Lab8 tables as data frames
plot_data <- bind_rows(
  summarize_table(raw.lm, "Clean"),
  summarize_table(sc.lm, "Suppression Corrected"),
  summarize_table(mstus.lm, "MSTUS Normalized"),
  summarize_table(norm.lm, "Clean Normalized")
)

plot_data$DataType <- factor(plot_data$DataType, 
                              levels = c("Clean", "Suppression Corrected", 
                                         "MSTUS Normalized", "Clean Normalized"))

plot_data$Term <- factor(plot_data$Term, levels = c("group", "concentration", "residuals"))

ggplot(plot_data, aes(x = DataType, y = Mean, fill = Term)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = paste0(round(Mean, 1), "% ± ", SD)),
            position = position_stack(vjust = 0.5), size = 3, color = "white") +
  scale_fill_manual(values = c("group" = "cornflowerblue", "concentration" = "darkolivegreen3", "residuals" = "#898781")) +
  labs(x = NULL, y = "Mean % contribution", fill = NULL,
       title = "Relative contribution to model by data type") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")

ggsave(ggsave("./outputs/lm_contributions_barchart.png", width = 12, height = 8))
```

#### Plot relative contributions per data type (faceted box plot)

```{r}
# Apply LM function to data types and summarize results
raw.lm.all <- lm.group.conc(lapply(raw.impute, log2), meta.data, summarize = F)
sc.lm.all <- lm.group.conc(lapply(sc.impute, log2), meta.data, summarize = F)
mstus.lm.all <- lm.group.conc(lapply(mstus.impute, log2), meta.data, summarize = F)
norm.lm.all <- lm.group.conc(lapply(norm.impute, log2), meta.data, summarize = F)

# Plot as faceted box plots

# Prepare data (concatenate data types)
raw.lm.all$data_type <- "Clean"
sc.lm.all$data_type <- "Suppression Corrected"
mstus.lm.all$data_type <- "MSTUS Normalized"
norm.lm.all$data_type <- "Clean Normalized"

all.lm.all <- rbind(raw.lm.all, sc.lm.all, mstus.lm.all, norm.lm.all)

# Prepare for ggplot

all.lm.plot <- all.lm.all %>%
    pivot_longer(
    cols = c(group, concentration, Residuals),
    names_to = "Term",
    values_to = "PctContribution"
  )

#  Set factor levels to control facet order
all.lm.plot$Term <- factor(all.lm.plot$Term,
                          levels = c("group", "concentration", "Residuals"))

# Set factor levels to control x-axis order
all.lm.plot$data_type <- factor(all.lm.plot$data_type,
                               levels = c("Clean", "Suppression Corrected",
                                          "MSTUS Normalized", "Clean Normalized"))

# Named color vector, keyed to the factor levels
term_colors <- c(
  "group" = "cornflowerblue",
  "concentration" = "darkolivegreen3",
  "Residuals" = "#898781"
)

ggplot(all.lm.plot, aes(x = data_type, y = PctContribution, fill = Term)) +
  geom_boxplot() +
  facet_wrap(~ Term, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = term_colors) +
  labs(x = "Data Type", y = "% Contribution") +
  theme_bw() +
  theme(legend.position = "none")

ggsave("./outputs/lm_contributions_boxplot.png", width = 12, height = 8)
```

### LM: PCA scores ~ group + concentration

```{r}
# Calculate PCAs
raw.pca <- lapply(raw.impute, function(x) make.pca(x, meta.data, "group", group.shape.name = "concentration", do.log2 = T, centered = T, scaled = T, graph = F))
sc.pca <- lapply(sc.impute, function(x) make.pca(x, meta.data, "group", group.shape.name = "concentration", do.log2 = T, centered = T, scaled = T, graph = F))
mstus.pca <- lapply(mstus.impute, function(x) make.pca(x, meta.data, "group", group.shape.name = "concentration", do.log2 = T, centered = T, scaled = T, graph = F))
norm.pca <- lapply(norm.impute, function(x) make.pca(x, meta.data, "group", group.shape.name = "concentration", do.log2 = T, centered = T, scaled = T, graph = F))

# Extract scores for PC1
raw.scores <- lapply(raw.pca, function(x) x$x[,c("PC1", "PC2", "PC3")])
sc.scores <- lapply(sc.pca, function(x) x$x[,c("PC1", "PC2", "PC3")])
mstus.scores <- lapply(mstus.pca, function(x) x$x[,c("PC1", "PC2", "PC3")])
norm.scores <- lapply(norm.pca, function(x) x$x[,c("PC1", "PC2", "PC3")])

# Apply linear model function
raw.lm.pc <- lm.group.conc(raw.scores, meta.data, summarize = F, pvals = T)
raw.lm.pc$data_type <- "Clean"

sc.lm.pc <- lm.group.conc(sc.scores, meta.data, summarize = F, pvals = T)
sc.lm.pc$data_type <- "Suppression Corrected"

mstus.lm.pc <- lm.group.conc(mstus.scores, meta.data, summarize = F, pvals = T)
mstus.lm.pc$data_type <- "MSTUS Normalized"

norm.lm.pc <- lm.group.conc(norm.scores, meta.data, summarize = F, pvals = T)
norm.lm.pc$data_type <- "Clean Normalized"

# Concatenate data for plotting
all.lm.pc <- rbind(raw.lm.pc, sc.lm.pc, mstus.lm.pc, norm.lm.pc)
```

#### Extract % explained variance for each PC

```{r}
raw.exp.var <- exp.var(raw.pca, n.pc = 3)
sc.exp.var <- exp.var(sc.pca, n.pc = 3)
mstus.exp.var <- exp.var(mstus.pca, n.pc = 3)
norm.exp.var <- exp.var(norm.pca, n.pc = 3)

total.var <- rbind(raw.exp.var, sc.exp.var, mstus.exp.var, norm.exp.var)

colMeans(total.var)

apply(total.var, 2, sd)
```


#### Summarize labs: plot explained variance and p-values for each data type 

##### PC1

```{r}
# --- Set consistent factor levels for x-axis order and colors ---
data_type_order <- c("Clean", "Suppression Corrected", "MSTUS Normalized", "Clean Normalized")

term_colors_pct <- c("group" = "cornflowerblue", "concentration" = "darkolivegreen3", "Residuals" = "darkgrey")
term_colors_pval <- c("group" = "cornflowerblue", "concentration" = "darkolivegreen3")

# --- Prep data: filter to PC1, reshape to long format for plotting ---

# Plot 1 data: PctContribution for group, concentration, Residuals
pct_long <- all.lm.pc %>%
  filter(Metabolite == "PC1") %>%
  select(data_type, PctContribution_group, PctContribution_concentration, PctContribution_Residuals) %>%
  pivot_longer(
    cols = starts_with("PctContribution_"),
    names_to = "Term",
    names_prefix = "PctContribution_",
    values_to = "PctContribution"
  ) %>%
  group_by(data_type, Term) %>%
  summarise(MeanPct = mean(PctContribution, na.rm = TRUE),
            SDPct = sd(PctContribution, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(
    Term = factor(Term, levels = c("group", "concentration", "Residuals")),
    data_type = factor(data_type, levels = data_type_order),
    Label = sprintf("%.1f \u00b1 %.1f", MeanPct, SDPct)
  ) %>%
  # KEY FIX: geom_col stacks in REVERSE factor level order (last level at bottom-most
  # drawn first, i.e. first level ends up on top). To get label y-positions that match
  # what's actually drawn, sort by the reverse of the factor order before cumsum().
  arrange(data_type, desc(Term)) %>%
  group_by(data_type) %>%
  mutate(
    StackTop = cumsum(MeanPct),
    LabelY = StackTop - MeanPct / 2
  ) %>%
  ungroup()

# Plot 2 data: -log10(p-value) for group and concentration only
pval_long <- all.lm.pc %>%
  filter(Metabolite == "PC1") %>%
  select(data_type, PValue_group, PValue_concentration) %>%
  pivot_longer(
    cols = starts_with("PValue_"),
    names_to = "Term",
    names_prefix = "PValue_",
    values_to = "PValue"
  ) %>%
  mutate(NegLogP = -log10(PValue)) %>%
  group_by(data_type, Term) %>%
  #summarise(MeanNegLogP = mean(NegLogP, na.rm = TRUE),
            #SDNegLogP = sd(NegLogP, na.rm = TRUE),
            #MeanP = mean(PValue, na.rm = TRUE),
            #.groups = "drop") %>%
  mutate(
    Term = factor(Term, levels = c("group", "concentration")),
    data_type = factor(data_type, levels = data_type_order),
  #  Label = sprintf("%.1f \u00b1 %.1f\n(p=%.2g)", MeanNegLogP, SDNegLogP, MeanP)
  )

# --- Plot 1: stacked bar of variance contribution ---
pct_long <- pct_long %>%
  mutate(x_num = as.numeric(factor(data_type)))

p1 <- ggplot(pct_long, aes(x = x_num, y = MeanPct, fill = Term)) +
  geom_col(aes(x = x_num - 0.1), position = "stack", width = 0.4) +
  geom_text(aes(x = x_num + 0.15, y = LabelY, label = Label), 
            color = "black", size = 3, hjust = 0) +
  facet_wrap(~ "PC1") +
  scale_fill_manual(values = term_colors_pct) +
  scale_x_continuous(breaks = unique(pct_long$x_num), 
                      labels = unique(pct_long$data_type)) +
  labs(y = "% Contribution to Variance", x = NULL, fill = "Term",
       title = "Variance Contribution by Term (PC1)") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        legend.position = "top") +
  coord_cartesian(xlim = c(0.5, length(unique(pct_long$x_num)) + 0.5))


# --- Plot 2: dodged bar of -log10(p-value) ---
p2 <- ggplot(pval_long, aes(x = data_type, y = NegLogP, fill = Term)) +
  geom_boxplot(position = position_dodge(width = 0.9)) +
 # geom_text(aes(label = Label), position = position_dodge(width = 0.9),
  #          vjust = -0.1, color = "black", size = 3) +
  facet_wrap(~ "PC1") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  scale_fill_manual(values = term_colors_pval) +
  labs(y = expression(-log[10](p*"-value")), x = "", fill = "Term",
       title = "Significance by Term (PC1)") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  coord_cartesian(clip = "off") 
# + expand_limits(y = max(pval_long$NegLogP + pval_long$SDNegLogP) * 1.15)

# --- Combine, stacked vertically ---
both.plot <- p1 / p2

ggsave("./outputs/lm_contributions_PC1_both.png", width = 12, height = 8)
```

##### PC2

```{r}
# --- Set consistent factor levels for x-axis order and colors ---
data_type_order <- c("Clean", "Suppression Corrected", "MSTUS Normalized", "Clean Normalized")

term_colors_pct <- c("group" = "cornflowerblue", "concentration" = "darkolivegreen3", "Residuals" = "darkgrey")
term_colors_pval <- c("group" = "cornflowerblue", "concentration" = "darkolivegreen3")

# --- Prep data: filter to PC1, reshape to long format for plotting ---

# Plot 1 data: PctContribution for group, concentration, Residuals
pct_long <- all.lm.pc %>%
  filter(Metabolite == "PC2") %>%
  select(data_type, PctContribution_group, PctContribution_concentration, PctContribution_Residuals) %>%
  pivot_longer(
    cols = starts_with("PctContribution_"),
    names_to = "Term",
    names_prefix = "PctContribution_",
    values_to = "PctContribution"
  ) %>%
  group_by(data_type, Term) %>%
  summarise(MeanPct = mean(PctContribution, na.rm = TRUE),
            SDPct = sd(PctContribution, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(
    Term = factor(Term, levels = c("group", "concentration", "Residuals")),
    data_type = factor(data_type, levels = data_type_order),
    Label = sprintf("%.1f \u00b1 %.1f", MeanPct, SDPct)
  ) %>%
  # KEY FIX: geom_col stacks in REVERSE factor level order (last level at bottom-most
  # drawn first, i.e. first level ends up on top). To get label y-positions that match
  # what's actually drawn, sort by the reverse of the factor order before cumsum().
  arrange(data_type, desc(Term)) %>%
  group_by(data_type) %>%
  mutate(
    StackTop = cumsum(MeanPct),
    LabelY = StackTop - MeanPct / 2
  ) %>%
  ungroup()

# Plot 2 data: -log10(p-value) for group and concentration only
pval_long <- all.lm.pc %>%
  filter(Metabolite == "PC2") %>%
  select(data_type, PValue_group, PValue_concentration) %>%
  pivot_longer(
    cols = starts_with("PValue_"),
    names_to = "Term",
    names_prefix = "PValue_",
    values_to = "PValue"
  ) %>%
  mutate(NegLogP = -log10(PValue)) %>%
  group_by(data_type, Term) %>%
  #summarise(MeanNegLogP = mean(NegLogP, na.rm = TRUE),
            #SDNegLogP = sd(NegLogP, na.rm = TRUE),
            #MeanP = mean(PValue, na.rm = TRUE),
            #.groups = "drop") %>%
  mutate(
    Term = factor(Term, levels = c("group", "concentration")),
    data_type = factor(data_type, levels = data_type_order),
  #  Label = sprintf("%.1f \u00b1 %.1f\n(p=%.2g)", MeanNegLogP, SDNegLogP, MeanP)
  )

# --- Plot 1: stacked bar of variance contribution ---
pct_long <- pct_long %>%
  mutate(x_num = as.numeric(factor(data_type)))

p1 <- ggplot(pct_long, aes(x = x_num, y = MeanPct, fill = Term)) +
  geom_col(aes(x = x_num - 0.1), position = "stack", width = 0.4) +
  geom_text(aes(x = x_num + 0.15, y = LabelY, label = Label), 
            color = "black", size = 3, hjust = 0) +
  facet_wrap(~ "PC2") +
  scale_fill_manual(values = term_colors_pct) +
  scale_x_continuous(breaks = unique(pct_long$x_num), 
                      labels = unique(pct_long$data_type)) +
  labs(y = "% Contribution to Variance", x = NULL, fill = "Term",
       title = "Variance Contribution by Term (PC2)") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        legend.position = "top") +
  coord_cartesian(xlim = c(0.5, length(unique(pct_long$x_num)) + 0.5))


# --- Plot 2: dodged bar of -log10(p-value) ---
p2 <- ggplot(pval_long, aes(x = data_type, y = NegLogP, fill = Term)) +
  geom_boxplot(position = position_dodge(width = 0.9)) +
 # geom_text(aes(label = Label), position = position_dodge(width = 0.9),
  #          vjust = -0.1, color = "black", size = 3) +
  facet_wrap(~ "PC2") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  scale_fill_manual(values = term_colors_pval) +
  labs(y = expression(-log[10](p*"-value")), x = "", fill = "Term",
       title = "Significance by Term (PC1)") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  coord_cartesian(clip = "off") 
# + expand_limits(y = max(pval_long$NegLogP + pval_long$SDNegLogP) * 1.15)

# --- Combine, stacked vertically ---
both.plot <- p1 / p2

p1
```

##### PC3

```{r}
# --- Set consistent factor levels for x-axis order and colors ---
data_type_order <- c("Clean", "Suppression Corrected", "MSTUS Normalized", "Clean Normalized")

term_colors_pct <- c("group" = "cornflowerblue", "concentration" = "darkolivegreen3", "Residuals" = "darkgrey")
term_colors_pval <- c("group" = "cornflowerblue", "concentration" = "darkolivegreen3")

# --- Prep data: filter to PC1, reshape to long format for plotting ---

# Plot 1 data: PctContribution for group, concentration, Residuals
pct_long <- all.lm.pc %>%
  filter(Metabolite == "PC3") %>%
  select(data_type, PctContribution_group, PctContribution_concentration, PctContribution_Residuals) %>%
  pivot_longer(
    cols = starts_with("PctContribution_"),
    names_to = "Term",
    names_prefix = "PctContribution_",
    values_to = "PctContribution"
  ) %>%
  group_by(data_type, Term) %>%
  summarise(MeanPct = mean(PctContribution, na.rm = TRUE),
            SDPct = sd(PctContribution, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(
    Term = factor(Term, levels = c("group", "concentration", "Residuals")),
    data_type = factor(data_type, levels = data_type_order),
    Label = sprintf("%.1f \u00b1 %.1f", MeanPct, SDPct)
  ) %>%
  # KEY FIX: geom_col stacks in REVERSE factor level order (last level at bottom-most
  # drawn first, i.e. first level ends up on top). To get label y-positions that match
  # what's actually drawn, sort by the reverse of the factor order before cumsum().
  arrange(data_type, desc(Term)) %>%
  group_by(data_type) %>%
  mutate(
    StackTop = cumsum(MeanPct),
    LabelY = StackTop - MeanPct / 2
  ) %>%
  ungroup()

# Plot 2 data: -log10(p-value) for group and concentration only
pval_long <- all.lm.pc %>%
  filter(Metabolite == "PC3") %>%
  select(data_type, PValue_group, PValue_concentration) %>%
  pivot_longer(
    cols = starts_with("PValue_"),
    names_to = "Term",
    names_prefix = "PValue_",
    values_to = "PValue"
  ) %>%
  mutate(NegLogP = -log10(PValue)) %>%
  group_by(data_type, Term) %>%
  #summarise(MeanNegLogP = mean(NegLogP, na.rm = TRUE),
            #SDNegLogP = sd(NegLogP, na.rm = TRUE),
            #MeanP = mean(PValue, na.rm = TRUE),
            #.groups = "drop") %>%
  mutate(
    Term = factor(Term, levels = c("group", "concentration")),
    data_type = factor(data_type, levels = data_type_order),
  #  Label = sprintf("%.1f \u00b1 %.1f\n(p=%.2g)", MeanNegLogP, SDNegLogP, MeanP)
  )

# --- Plot 1: stacked bar of variance contribution ---
pct_long <- pct_long %>%
  mutate(x_num = as.numeric(factor(data_type)))

p1 <- ggplot(pct_long, aes(x = x_num, y = MeanPct, fill = Term)) +
  geom_col(aes(x = x_num - 0.1), position = "stack", width = 0.4) +
  geom_text(aes(x = x_num + 0.15, y = LabelY, label = Label), 
            color = "black", size = 3, hjust = 0) +
  facet_wrap(~ "PC3") +
  scale_fill_manual(values = term_colors_pct) +
  scale_x_continuous(breaks = unique(pct_long$x_num), 
                      labels = unique(pct_long$data_type)) +
  labs(y = "% Contribution to Variance", x = NULL, fill = "Term",
       title = "Variance Contribution by Term (PC3)") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        legend.position = "top") +
  coord_cartesian(xlim = c(0.5, length(unique(pct_long$x_num)) + 0.5))


# --- Plot 2: dodged bar of -log10(p-value) ---
p2 <- ggplot(pval_long, aes(x = data_type, y = NegLogP, fill = Term)) +
  geom_boxplot(position = position_dodge(width = 0.9)) +
 # geom_text(aes(label = Label), position = position_dodge(width = 0.9),
  #          vjust = -0.1, color = "black", size = 3) +
  facet_wrap(~ "PC3") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  scale_fill_manual(values = term_colors_pval) +
  labs(y = expression(-log[10](p*"-value")), x = "", fill = "Term",
       title = "Significance by Term (PC3)") +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  coord_cartesian(clip = "off") 
# + expand_limits(y = max(pval_long$NegLogP + pval_long$SDNegLogP) * 1.15)

# --- Combine, stacked vertically ---
both.plot <- p1 / p2

p1
```