---
title: "20250717_Round Robin_RandomForest"
author: "Haley"
date: "2025-07-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code") 
```

# Unsupervised Random Forest clustering of Round Robin labs

This file performs unsupervised RF clustering, based on sample groups, for each lab and creates plots of sample clusters

## Load Functions and Data

```{r}
# Install/load randomForest package
if(!require("randomForest")){
  install.packages("randomForest")
  library(randomForest())
}

if(!require("cluster")){
  install.packages("cluster")
  library(cluster)
}

source("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/code/plotting_functions.R")
source("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/code/reproducibility_functions.R")
source("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/code/summary_functions.R")

load("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/reproducibility_input/duplicated_filtered_data.RData")
```

## Run Unsupervised RF on each lab

```{r}
norm.log <- lapply(norm.std, log2)

set.seed(333)
rfs <- lapply(norm.log, function(x) randomForest(x = x, proximity = T, replace = F))

```

### Check removing one Lab 3 duplicate

```{r}
norm.log <- lapply(norm.std, log2)

norm.log.3fix <- norm.log

norm.log.3fix$Lab3 <- norm.log.3fix$Lab3[-which(row.names(norm.log.3fix$Lab3) == "ab_0.3_rep2"),]

set.seed(333)
rfs.3fix <- lapply(norm.log.3fix, function(x) randomForest(x = x, proximity = T, replace = F))
```

## Plot clusters for each lab using unsupervised random forest

```{r}
# set group colors
group.colors <- c("red","orange","brown","purple","gold","darkgreen", "blue")
names(group.colors) <- levels(meta.data$group)

group.colors.df <- data.frame(group = names(group.colors), group.color = group.colors)

# set concentration colors
conc.colors <- brewer.pal(length(unique(meta.data$concentration)), "Blues")
names(conc.colors) <- levels(meta.data$concentration)

conc.colors.df <- data.frame(concentration = names(conc.colors), conc.color = conc.colors)

# add colors to meta data
meta.data.new <- merge(meta.data, group.colors.df, by = "group")
meta.data.new <- merge(meta.data.new, conc.colors.df, by = "concentration")
meta.data.new <- meta.data.new[order(meta.data.new$group),]
row.names(meta.data.new) <- row.names(meta.data)

# build distance metric
dists <- lapply(rfs, function(x) dist(1 - x$proximity))

# input for multidimensional scaling in 2D using mds
mds <- lapply(dists, function(x) cmdscale(x, k = 2)) # try k-means to get cluster differences
mds.group.colors <- lapply(mds, function(x) cbind(x, meta.data.new$group.color))

# plot mds
png("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/unsupervisedRF_clusters_WITHdupes.png", width = 8, height = 4, units = "in", res = 600)
par(mfrow = c(2, 4), mgp = c(1.5,0.5,0), mar = c(3,3,2,2))
mapply(function (x, y) plot(x, pch=16, xlab="Dim1", ylab="Dim2", col = meta.data.new$group.color, cex = 1, main=paste(y)), 
       mds.group.colors, 
       names(mds.group.colors))
```

# calculate strength of clustering using k-means cluster assignments and silhouette

```{r}
# check how k-means clustering performs on 2D mds data from all labs

for(i in 1:length(mds)){
  
  k.means.i <- kmeans(mds[[i]], centers = length(unique(meta.data$group)))
  
  plot(mds.group.colors[[i]], pch=16, xlab="Dim1", ylab="Dim2", col = meta.data.new$group.color, cex = 2, main=paste(names(mds)[[i]]))
points(k.means.i$centers[,1], k.means.i$centers[,2], pch = 19)
  
}

# calculate silhouette scores based on a priori clusters (from meta-data) and plot


for(i in 1:length(mds)){
  
  sil.i <- silhouette(as.numeric(meta.data$group), dist(mds[[i]]))
  
  print(fviz_silhouette(sil.i))
  
}

```

## Check removing one sample in Lab 3

```{r}
# build distance metric
dists.3fix <- lapply(rfs.3fix, function(x) dist(1 - x$proximity))

# input for multidimensional scaling in 2D using mds
mds.3fix <- lapply(dists.3fix, function(x) cmdscale(x, k = 2)) # try k-means to get cluster differences
mds.group.colors.3fix <- lapply(mds.3fix, function(x) cbind(x, meta.data.new$group.color))

# check how k-means clustering performs on 2D mds data from all labs

for(i in 1:length(mds.3fix)){
  
  k.means.i <- kmeans(mds.3fix[[i]], centers = length(unique(meta.data$group)))
  
  plot(mds.group.colors[[i]], pch=16, xlab="Dim1", ylab="Dim2", col = meta.data.new$group.color, cex = 2, main=paste(names(mds)[[i]]))
points(k.means.i$centers[,1], k.means.i$centers[,2], pch = 19)
  
}

# calculate silhouette scores based on a priori clusters (from meta-data) and plot


for(i in 1:length(mds.3fix)){
  
  if(i != 3){
    
  sil.i <- silhouette(as.numeric(meta.data$group), dist(mds.3fix[[i]]))
  
  print(fviz_silhouette(sil.i))
    
  } else {
    
   # meta.data fix

  meta.data.3fix <- meta.data[-which(row.names(meta.data) == "ab_0.3_rep2"),] 
    
  sil.i <- silhouette(as.numeric(meta.data.3fix$group), dist(mds.3fix[[i]]))
  
  print(fviz_silhouette(sil.i))
  
  }
  

  
}

```
