---
title: "06_NMF_clean"
output: html_document
date: "2026-07-16"
---

# Load packages and functions

```{r}
source("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/code/nmf_functions.R")
```

# Load data

```{r}
load("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/duplicate_input/imputed_data.RData")

#par(mfrow = c(2,4))
lapply(norm.impute, function(x) plot(density(apply(x, 2, as.numeric)), main = ""))

#par(mfrow = c(2, 4))
lapply(norm.impute, function(x) boxplot((apply(x, 2, as.numeric)), main = ""))
```

# Test different values for rank

```{r}
ranks <- c(2, 3, 4, 5, 6)

true.ranks <- nmf_rank(norm.impute, rank.range = ranks, nrun = 30, randomize = FALSE, plot.list = c("evar", "rss", "silhouette"))
rando.ranks <- nmf_rank(norm.impute, rank.range = ranks, nrun = 30, randomize = TRUE, plot.list = c("evar", "rss", "silhouette"))

grid.arrange(grobs = true.ranks, ncol = 4, top = "Real data")
grid.arrange(grobs = rando.ranks, ncol = 4, top = "Randomized data")
```

# Report NMF performance with rank = 3

```{r}
all.perf <- nmf_performance(norm.impute, rank = 3, seed = 3, meta.data)

# see if evar() has same results
all.var <- nmf_evar(norm.impute, rank = 3, seed = 3, meta.data)
```

# Run NMF, plot heatmaps, and calculate performance of clustering 

```{r}
png("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/NMF_heatmap_WITH_dupes.png", width = 36, height = 18, units = "in", res = 600)

w <- nmf_basis(norm.impute, 3, meta.data, "group", print.heatmap = T, seed = 3)

png("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/NMF_heatmap_WITH_dupes_w_conc.png", width = 40, height = 30, units = "in", res = 600)

w <- nmf_basis(norm.impute, 3, meta.data, "both", print.heatmap = T, seed = 3)

```

# Validate that clusters are according to sample group

```{r}
nmf_cluster(w$w, meta.data)
ggsave("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/NMF_cluster_correlations_WITH_dupes.png", width = 12, height = 8, dpi = 600)

```

# Calculate ARI as another validation metric

```{r}
ari <- ari.nmf(w, meta.data)
```

# Determine important features

```{r}
h <- nmf_feats(norm.impute, rank = 3, meta.data, meta.col = "group", comp.found.na.filter)
```

## save list of features for biomarker filtering

```{r}
save(h, comp.found.na.filter, file = "/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/biomarker_input/NMF_features_WITH_dupes.RData")
```

# Calculate feature scores

```{r}
# calculate score (contribution of feature to basis)
scores <- lapply(h, function(x) featureScore(t(x), method = "kim"))

# graph distribution of scores
lapply(scores, function(x) hist(x, breaks = 100))
```

## save list of feature scores for biomarker filtering

```{r}
save(scores, comp.found.na.filter, file = "/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/biomarker_input/NMF_feature_scores_WITH_dupes.RData")
```

# Permutation test

```{r}
# Run Test
#w.1000 <- permute.w(norm.impute, 1000, meta.data)

save(w.1000, comp.found.na.filter, file = "/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/biomarker_input/NMF_1000_iterations.RData")

load("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/biomarker_input/NMF_1000_iterations.RData")

# Report significance
ari.p <- perm.sig(w.1000, ari, "ARI")
corr.p <- perm.sig(w.1000, w$correlation, "Correlation")

# Report summary of permutation results
ari.results <- permute.sum(w.1000, "ARI")
corr.results <- permute.sum(w.1000, "Correlation")

```

# Train/Test test

```{r}
# Correct split
true.train.test <- nmf.tt(norm.impute, 
                          split.reps = T,
                          split.n = NULL, 
                          rank = 3,
                          sample.seed = NULL,
                          nmf.seed = 3, 
                          meta.data)

# Random splits
iterations <- seq_len(10)

randos.train.test <- lapply(iterations, function(x){
  cat("iteration:", x, "\n")
  nmf.tt(norm.impute, 
                                                           split.reps = F, 
                                                           split.n = 16, 
                                                           rank = 3, 
                                                           sample.seed = x, 
                                                           nmf.seed = 3, 
                                                           meta.data)
} )

# Report results

accuracy_mat <- sapply(randos.train.test, function(iter_result) {
  sapply(iter_result, function(lab_df) sum(lab_df$correct) / nrow(lab_df) * 100)
})
# accuracy_mat: rows = labs, columns = iterations

accuracy_summary <- data.frame(
  lab = rownames(accuracy_mat),
  mean_pct_correct = apply(accuracy_mat, 1, mean),
  sd_pct_correct = apply(accuracy_mat, 1, sd)
)

accuracy_summary
```