---
title: "20250124_RoundRobin_BiomarkerFiltering"
author: "Haley"
date: "2025-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load data

```{r}
load("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/biomarker_input/ANOVA_features_WITH_duplicates.RData")
load("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/biomarker_input/PCA_features_WITHduplicates.RData")
load("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/biomarker_input/NMF_features_WITH_dupes.RData")
```

# Load packages

```{r}
packages <- c("tidyverse", "dplyr", "ggplot2", "ggpattern", "plotly", 
              "RColorBrewer", "ggsci", "ggpubr", "pheatmap", "grafify", "patchwork",
              "openxlsx", "data.table", "grafify", "factoextra", "htmltools", "colorspace")

installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

invisible(lapply(packages, library, character.only = TRUE))

BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
```

# ANOVA features filter
features with adjusted p-values < 0.05

```{r}
anova.feats <- list(length(anova.results))

anova.feats <- lapply(anova.results, function(x) which(x$`adj.p` < 0.05))

anova.feats <- mapply(function(x, y) x$feature[y], anova.results, anova.feats)
```

# PCA features filter
top 10% absolute value loadings from each PC (take unique across all 3 PCs)

```{r}
pca.feats <- lapply(norm.group.load, abs)

pca.filter <- function(lab.feats, cutoff){
  # lab.feats = pca features from one lab
  # cutoff = percentile cutoff
  
  results <- apply(lab.feats, 2, function(x) which(x > quantile(x, cutoff)))
  
  results <- apply(results, 2, function(x) row.names(lab.feats)[x])
  
  return(results)

}

pca.feats <- lapply(pca.feats, function(x) pca.filter(x, 0.9))

pca.feats <- lapply(pca.feats, function(x) unique(c(x[,1], x[,2], x[,3])) )
```

Find the quantiles for GPC

```{r}
gpc <- list("Lab1" = c("LTRS_000076_SECIM-RP-Pos"),
            "Lab2" = c("LTRS_000149_BasePos", "LTRS_000044_BaseNeg"),
            "Lab3" = c("LTRS_000112_MDARP-pos", "LTRS_000131_MDARP-pos"),
            "Lab4" = c("LTRS_000040_BarnesPos"),
            "Lab5" = c("LTRS_000041_BasePos", "LTRS_000076_BasePos"),
            "Lab6" = c("LTRS_000060_BasePos"),
            "Lab7" = c("LTRS_000106_BasePos", "LTRS_000114_BasePos"),
            "Lab8" = c("LTRS_000143_BasePos", "LTRS_000128_BasePos"))

pca.test <- lapply(norm.group.load, abs)

for(lab in 1:length(pca.test)){
  for(pc in 1:ncol(pca.test[[lab]])){
  
  if(length(gpc[[lab]]) == 1){
    
   pc_ecdf <- ecdf(pca.test[[lab]][,pc]) 
   
   print(paste("Qunatile of", gpc[[lab]], "in lab", lab, "in PC", pc, "=", pc_ecdf(pca.test[[lab]][which(row.names(pca.test[[lab]]) == gpc[[lab]]), pc])))
   
  } else {
  
    for(feat in 1:length(gpc[[lab]])){
      
   pc_ecdf <- ecdf(pca.test[[lab]][,pc]) 
   
   print(paste("Qunatile of", gpc[[lab]][feat], "in lab", lab, "in PC", pc, "=", pc_ecdf(pca.test[[lab]][which(row.names(pca.test[[lab]]) == gpc[[lab]][feat]), pc])))      
      
      }
    }
  }
}
```


# NMF feature filter
features that contribute at least two-thirds to a given rank

```{r}
nmf_feat_imp <- function(h, imp.cutoff = 0.67){
  # h = matrix of feature coefficients expressed as a fraction of total contribution to each rank
  # imp.cutoff = value for feature importance within each rank
  
  # determine features with percent coefficient values above cutoff for each rank
    h.feats <- list(length(h))
    
    for(i in 1:length(h)){
      
      a.feats <- colnames(h[[i]])[which(h[[i]][c("a"),] > imp.cutoff)]
      b.feats <- colnames(h[[i]])[which(h[[i]][c("b"),] > imp.cutoff)]
      c.feats <- colnames(h[[i]])[which(h[[i]][c("c"),] > imp.cutoff)]
      
      h.feats[[i]] <- list("a" = a.feats, "b" = b.feats, "c" = c.feats)
      
    }
    
    names(h.feats) <- names(h)
  
  return(h.feats)
  
}

nmf.feats <- nmf_feat_imp(h, 0.67)

```

# Table of features and importance

```{r}
all.feats <- c(unlist(anova.feats), unlist(pca.feats), unlist(nmf.feats)) 

results <- list()

for(i in 1:8){ # for all labs
  print(i)
  
  results[[i]] <- data.frame(matrix(ncol = 7))
  colnames(results[[i]]) <- c("Name", "lab", "MF", "InChiKey", "ANOVA", "PCA", "NMF")
  
  for(j in 1:length(all.feats)){ # for all features that are significant anywhere
    #print(j)
    
    # if in all
    
    if(all.feats[j] %in% anova.feats[[i]] &
       all.feats[j] %in% pca.feats[[i]] &
       all.feats[j] %in% unlist(nmf.feats[[i]])){
      
      results[[i]] <- rbind(results[[i]], c("Name" = all.feats[j], 
                                            "lab" = names(anova.feats)[[i]],
                                            "MF" = unique(comp.found.na.filter[[i]]$MF[
                                              which(row.names(comp.found.na.filter[[i]]) == all.feats[j])]),
                                            "InChiKey" = comp.found.na.filter[[i]]$InChiKey2D[which(row.names(comp.found.na.filter[[i]]) == all.feats[j])],
                                            "ANOVA" = 1, 
                                            "PCA" = 1, 
                                            "NMF" = 1))
    }
    
    # if in ANOVA and PCA
    
    else if(all.feats[j] %in% anova.feats[[i]] &
       all.feats[j] %in% pca.feats[[i]] &
       all.feats[j] %in% unlist(nmf.feats[[i]]) == F){
      
      results[[i]] <- rbind(results[[i]], c("Name" = all.feats[j], 
                                            "lab" = names(anova.feats)[[i]],
                                            "MF" = unique(comp.found.na.filter[[i]]$MF[
                                              which(row.names(comp.found.na.filter[[i]]) == all.feats[j])]),
                                            "InChiKey" = comp.found.na.filter[[i]]$InChiKey2D[which(row.names(comp.found.na.filter[[i]]) == all.feats[j])],
                                            "ANOVA" = 1, 
                                            "PCA" = 1, 
                                            "NMF" = 0))
        }    
    
    # if in ANOVA and NMF
    
    else if(all.feats[j] %in% anova.feats[[i]] &
       all.feats[j] %in% pca.feats[[i]] == F &
       all.feats[j] %in% unlist(nmf.feats[[i]])){
      
      results[[i]] <- rbind(results[[i]], c("Name" = all.feats[j], 
                                            "lab" = names(anova.feats)[[i]],
                                            "MF" = unique(comp.found.na.filter[[i]]$MF[
                                              which(row.names(comp.found.na.filter[[i]]) == all.feats[j])]),
                                            "InChiKey" = comp.found.na.filter[[i]]$InChiKey2D[which(row.names(comp.found.na.filter[[i]]) == all.feats[j])],
                                            "ANOVA" = 1, 
                                            "PCA" = 0, 
                                            "NMF" = 1))
        }      
 
    # if in PCA and NMF
    
    else if(all.feats[j] %in% anova.feats[[i]] == F &
       all.feats[j] %in% pca.feats[[i]] &
       all.feats[j] %in% unlist(nmf.feats[[i]])){
      
      results[[i]] <- rbind(results[[i]], c("Name" = all.feats[j], 
                                            "lab" = names(anova.feats)[[i]],
                                            "MF" = unique(comp.found.na.filter[[i]]$MF[
                                              which(row.names(comp.found.na.filter[[i]]) == all.feats[j])]),
                                            "InChiKey" = comp.found.na.filter[[i]]$InChiKey2D[which(row.names(comp.found.na.filter[[i]]) == all.feats[j])],
                                            "ANOVA" = 0, 
                                            "PCA" = 1, 
                                            "NMF" = 1))
        }        
  
    # if in only ANOVA
    
    else if(all.feats[j] %in% anova.feats[[i]] &
       all.feats[j] %in% pca.feats[[i]] == F &
       all.feats[j] %in% unlist(nmf.feats[[i]]) == F){
      
      results[[i]] <- rbind(results[[i]], c("Name" = all.feats[j], 
                                            "lab" = names(anova.feats)[[i]],
                                            "MF" = unique(comp.found.na.filter[[i]]$MF[
                                              which(row.names(comp.found.na.filter[[i]]) == all.feats[j])]),
                                            "InChiKey" = comp.found.na.filter[[i]]$InChiKey2D[which(row.names(comp.found.na.filter[[i]]) == all.feats[j])],
                                            "ANOVA" = 1, 
                                            "PCA" = 0, 
                                            "NMF" = 0))
        }     
     
    # if in only PCA
    
    else if(all.feats[j] %in% anova.feats[[i]] == F &
       all.feats[j] %in% pca.feats[[i]] &
       all.feats[j] %in% unlist(nmf.feats[[i]]) == F){
      
      results[[i]] <- rbind(results[[i]], c("Name" = all.feats[j], 
                                            "lab" = names(anova.feats)[[i]],
                                            "MF" = unique(comp.found.na.filter[[i]]$MF[
                                              which(row.names(comp.found.na.filter[[i]]) == all.feats[j])]),
                                            "InChiKey" = comp.found.na.filter[[i]]$InChiKey2D[which(row.names(comp.found.na.filter[[i]]) == all.feats[j])],
                                            "ANOVA" = 0, 
                                            "PCA" = 1, 
                                            "NMF" = 0))
    } 
    
    # if in only NMF
    
    else if(all.feats[j] %in% anova.feats[[i]] == F &
       all.feats[j] %in% pca.feats[[i]] == F &
       all.feats[j] %in% unlist(nmf.feats[[i]])){
      
      results[[i]] <- rbind(results[[i]], c("Name" = all.feats[j], 
                                            "lab" = names(anova.feats)[[i]],
                                            "MF" = unique(comp.found.na.filter[[i]]$MF[
                                              which(row.names(comp.found.na.filter[[i]]) == all.feats[j])]),
                                            "InChiKey" = comp.found.na.filter[[i]]$InChiKey2D[which(row.names(comp.found.na.filter[[i]]) == all.feats[j])],
                                            "ANOVA" = 0, 
                                            "PCA" = 0, 
                                            "NMF" = 1))
        }        
  }
  
  results[[i]] <- results[[i]][-which(rowSums(is.na(results[[i]])) == ncol(results[[i]])),]
}

results.df <- data.frame()

for(k in 1:length(results)){
  
  results.df <- rbind(results.df, results[[k]])
  
}

# add number of analyses where the metabolite was important

sig.numbers <- apply(subset(results.df, select = c("ANOVA", "PCA", "NMF")), 2, as.numeric)
results.df$total.sig <- apply(sig.numbers, 1, sum)

# add number of labs where the MF was important in at least one analysis
mf <- unique(results.df$MF)

results.df$total.lab.mf <- NA

for(i in 1:length(mf)){
  
  subset.data <- results.df[which(results.df$MF == mf[i]),]
  
  n.labs <- length(unique(subset.data$lab))
 
  for(j in 1:nrow(results.df)){
    
    if(results.df$MF[j] == mf[i]){
      
      results.df$total.lab.mf[j] <- n.labs
      
    }
    
  }
   
}

# add number of labs where the InChiKey was important in at least one analysis
inchi <- unique(results.df$InChiKey)

results.df$total.lab.inchi <- NA

for(i in 1:length(inchi)){
  
  if(is.na(inchi[i])){next}
  
  subset.data <- results.df[which(results.df$InChiKey == inchi[i]),]
  
  n.labs <- length(unique(subset.data$lab))
 
  for(j in 1:nrow(results.df)){
    
    if(is.na(results.df$InChiKey[j])){next}
    
    if(results.df$InChiKey[j] == inchi[i]){
      
      results.df$total.lab.inchi[j] <- n.labs
      
    }
    
  }
   
}

write.csv(results.df, file = "/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/biomarker_input/important_features_WITH_duplicates.csv", row.names = F)
```

## How many metabolites (based on InChiKey or MF) were significant in at least one analysis in X labs?

```{r}
n.sig.labs <- data.frame(matrix(nrow = length(unique(results.df$total.lab.mf)), ncol = 3))
colnames(n.sig.labs) <- c("Number of Labs", "Number of Metabs by MF", "Number of Metabs by InChiKey")

for(i in 1:nrow(n.sig.labs)){
  
  n.sig.labs$`Number of Labs`[i] <- paste0("Significant in ", i, " labs")
  
  n.sig.labs$`Number of Metabs by MF`[i] <- length(unique(results.df$MF[which(results.df$total.lab.mf == i 
                                                        # & results.df$total.sig == 3
                                                                              )]))
  
  n.sig.labs$`Number of Metabs by InChiKey`[i] <- length(unique(results.df$InChiKey[which(results.df$total.lab.inchi == i 
                                                        #  & results.df$total.sig == 3
                                                                                          )]))
  
}

n.sig.analyses <- data.frame(matrix(nrow = length(unique(results.df$total.sig)), ncol = 3))
colnames(n.sig.analyses) <- c("Number of analyses", "Number of Metabs")

for(i in 1:nrow(n.sig.analyses)){
  
  n.sig.analyses$`Number of analyses`[i] <- paste0("Significant in ", i, " analyses")
  
  n.sig.analyses$`Number of Metabs by MF`[i] <- length(unique(results.df$MF[which(results.df$total.sig == i)]))
  
  n.sig.analyses$`Number of Metabs by InChiKey`[i] <- length(unique(results.df$InChiKey[which(results.df$total.sig == i)]))
  
}

```

# Prep data: sort by MF or InChiKey

## prep for re-organizing by MF

```{r}

load("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/duplicate_input/imputed_data.RData")
load("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/data/reproducibility_input/duplicated_filtered_data.RData")

metabs.8 <- unique(results.df$MF[which(results.df$total.lab.mf == 8 
                                                   #& results.df$total.sig == 3
                                                  )])

# Subset and prep abundance data

subset.metabs <- function(data.list, metab.list, lab, comp.meta, column, scale.metabs = F){
                      # data.list = metabolite abundance data in all labs
                      # metab.list = list of metabs to subset into
                      # lab = numeric index of lab
                      # comp.meta = metabolite meta data for all labs, no duplicate filtering
                      # column = column used to subset (MF or InChiKey)
                      # scale.metabs = logical indicating whether to scale abundances for each lab (recommended for heatmap)
                      
                      
                      lab.comp <- comp.meta[[lab]]
                      lab.data <- data.list[[lab]]
                      colnames(lab.data) <- row.names(lab.comp)
                      
                      subset.data <- lab.data[,which(lab.comp[,c(column)] %in% metab.list)]
                      
                      if(scale.metabs == T){
                        
                       subset.data <- scale(subset.data, scale = T, center = T)
                         
                      }else{
                        
                        subset.data <- subset.data
                        
                      }
                      
                      subset.comp <- subset(lab.comp[which(lab.comp[,c(column)] %in% metab.list),],
                                            select = c(Name, MF, InChiKey2D))
                      
                      return(list(data = subset.data, meta = subset.comp))
                      
                    }

lab.subset.data<- lapply(c("Lab1", "Lab2", "Lab3", "Lab4", "Lab5", "Lab6", "Lab7", "Lab8"), 
                          function(x) subset.metabs(norm.impute, metabs.8, x, comp.found.na.filter, "MF", scale.metabs = T))
names(lab.subset.data) <- names(norm.impute)
```

# Reorganize data to be subset by MF (not by lab) and only include features that were significant in all three analyses

```{r}
# list significant features
feats.sig <- c(unlist(anova.feats), unlist(pca.feats), unlist(nmf.feats)) 

feats.sig <- unique(feats.sig)

# create empty results list
mf.subset <- list()
length(mf.subset) <- length(metabs.8)
names(mf.subset) <- metabs.8

# loop through results to subset and reorganize

    for(mf in 1:length(mf.subset)){
      for(lab in 1:length(lab.subset.data)){
        
       # find column indexes in lab.subset.data with MF == mf
        cols <- which(lab.subset.data[[lab]]$meta$MF == names(mf.subset)[[mf]])
        
      # subset and melt data
      
      if(length(cols) > 1){
        
        new.data <- lab.subset.data[[lab]]$data[,cols]

        melt.data <- reshape2::melt(new.data)
        
        colnames(melt.data) <- c("sample.name", "feature", "value")
        
      # add sample group and lab name columns
        melt.data$lab <- names(lab.subset.data)[[lab]]
        
        melt.data$sample.group <- gsub("_.*", "", melt.data$sample.name)
        
      # add MF column
        
        mfs <- rownames_to_column(lab.subset.data[[lab]]$meta, var = "feature")
        
        fin.data <- merge(melt.data, subset(mfs, select = c(feature, MF, Name)), by = "feature")
        
      # add column indicating if a feature was also selected as a "Most Likely Feature" during duplicate filtering
        
        fin.data$retained <- NA
        
        for(row in 1:nrow(fin.data)){
          
          if(fin.data$feature[row] %in% row.names(comp.std[[lab]])){
            
            fin.data$retained[row] <- "X"
            
          }
          
        }
        
        fin.data$retained[is.na(fin.data$retained)] <- ""
        
      # only retain significant feature bins
        fin.data <- fin.data[which(fin.data$feature %in% feats.sig),]
  
      # append lab name to feature bin to avoid duplicate confusion
        fin.data$feature <- paste0(fin.data$lab, "_", fin.data$feature)
        
      }else{
        
        # subset data
        melt.data <- data.frame("sample.name" = row.names(lab.subset.data[[lab]]$data),
                               "feature" = row.names(lab.subset.data[[lab]]$meta)[cols],
                               "value" = lab.subset.data[[lab]]$data[,cols])
        
        # add sample group, MF, and lab name columns
        melt.data$lab <- names(lab.subset.data)[[lab]]
        
        melt.data$sample.group <- gsub("_.*", "", melt.data$sample.name)
        
        # add MF and feature name column
        
        mfs <- rownames_to_column(lab.subset.data[[lab]]$meta, var = "feature")
        
        fin.data <- merge(melt.data, subset(mfs, select = c(feature, MF, Name)), by = "feature")
        
              # add column indicating if a feature was also selected as a "Most Likely Feature" during duplicate filtering
        
        fin.data$retained <- NA
        
        for(row in 1:nrow(fin.data)){
          
          if(fin.data$feature[row] %in% row.names(comp.std[[lab]])){
            
            fin.data$retained[row] <- "X"
            
          }
          
        }
        
        fin.data$retained[is.na(fin.data$retained)] <- ""
        
        # only retain significant feature bins
        fin.data <- fin.data[which(fin.data$feature %in% feats.sig),]
        
        # append lab name to feature bin to avoid duplicate confusion
        fin.data$feature <- paste0(fin.data$lab, "_", fin.data$feature)
        
      }
       
        
      # add to results list
        if(lab == 1){
          
          mf.subset[[mf]] <- fin.data
          
        }else{
          
          mf.subset[[mf]] <- rbind(mf.subset[[mf]], fin.data)
          
        }
      }
    }
```

# Reorganize to subset by InChiKey

```{r}
# Reorganize data to be subset by InChiKey (not by lab) and only include features that were significant in at least one analysis

# subset by lab
metabs.8.inchi <- unique(results.df$InChiKey[which(results.df$total.lab.inchi == 8 
                                                  # & results.df$total.sig == 3
                                                  )])

lab.subset.data.inchi <- lapply(c("Lab1", "Lab2", "Lab3", "Lab4", "Lab5", "Lab6", "Lab7", "Lab8"), 
                          function(x) subset.metabs(norm.impute, metabs.8.inchi, x, comp.found.na.filter, "InChiKey2D", scale.metabs = T))
names(lab.subset.data.inchi) <- names(norm.impute)

# list significant features
feats.sig <- c(unlist(anova.feats), unlist(pca.feats), unlist(nmf.feats)) 

feats.sig <- unique(feats.sig)

# create empty results list
inchi.subset <- list()
length(inchi.subset) <- length(metabs.8.inchi)
names(inchi.subset) <- metabs.8.inchi

# loop through results to subset and reorganize

    for(inchi in 1:length(inchi.subset)){
      for(lab in 1:length(lab.subset.data.inchi)){
        
       # find column indexes in lab.subset.data.inchi with inchi == inchi
        cols <- which(lab.subset.data.inchi[[lab]]$meta$InChiKey2D == names(inchi.subset)[[inchi]])
        
      # subset and melt data
      
      if(length(cols) > 1){
        
        new.data <- lab.subset.data.inchi[[lab]]$data[,cols]

        melt.data <- reshape2::melt(new.data)
        
        colnames(melt.data) <- c("sample.name", "feature", "value")
        
      # add sample group and lab name columns
        melt.data$lab <- names(lab.subset.data.inchi)[[lab]]
        
        melt.data$sample.group <- gsub("_.*", "", melt.data$sample.name)
        
      # add InChi and MF columns
        
        inchis <- rownames_to_column(lab.subset.data.inchi[[lab]]$meta, var = "feature")
        
        fin.data <- merge(melt.data, subset(inchis, select = c(feature, InChiKey2D, Name, MF)), by = "feature")
        
      # add column indicating if a feature was also selected as a "Most Likely Feature" during duplicate filtering
        
        fin.data$retained <- NA
        
        for(row in 1:nrow(fin.data)){
          
          if(fin.data$feature[row] %in% row.names(comp.std[[lab]])){
            
            fin.data$retained[row] <- "X"
            
          }
          
        }
        
        fin.data$retained[is.na(fin.data$retained)] <- ""
        
      # only retain significant feature bins
        fin.data <- fin.data[which(fin.data$feature %in% feats.sig),]
  
      # append lab name to feature bin to avoid duplicate confusion
        fin.data$feature <- paste0(fin.data$lab, "_", fin.data$feature)
        
      }else{
        
        # subset data
        melt.data <- data.frame("sample.name" = row.names(lab.subset.data.inchi[[lab]]$data),
                               "feature" = row.names(lab.subset.data.inchi[[lab]]$meta)[cols],
                               "value" = lab.subset.data.inchi[[lab]]$data[,cols])
        
        # add sample group and lab name columns
        melt.data$lab <- names(lab.subset.data.inchi)[[lab]]
        
        melt.data$sample.group <- gsub("_.*", "", melt.data$sample.name)
        
      # add InChi and MF columns
        
        inchis <- rownames_to_column(lab.subset.data.inchi[[lab]]$meta, var = "feature")
        
        fin.data <- merge(melt.data, subset(inchis, select = c(feature, InChiKey2D, Name, MF)), by = "feature")
        
      # add column indicating if a feature was also selected as a "Most Likely Feature" during duplicate filtering
        
        fin.data$retained <- NA
        
        for(row in 1:nrow(fin.data)){
          
          if(fin.data$feature[row] %in% row.names(comp.std[[lab]])){
            
            fin.data$retained[row] <- "X"
            
          }
          
        }
        
        fin.data$retained[is.na(fin.data$retained)] <- ""
        
      # only retain significant feature bins
        fin.data <- fin.data[which(fin.data$feature %in% feats.sig),]
  
      # append lab name to feature bin to avoid duplicate confusion
        fin.data$feature <- paste0(fin.data$lab, "_", fin.data$feature)
        
      }
        
      # add to results list
        if(lab == 1){
          
          inchi.subset[[inchi]] <- fin.data
          
        }else{
          
          inchi.subset[[inchi]] <- rbind(inchi.subset[[inchi]], fin.data)
          
        }
      }
    }

```

# Remove peptides from the metabolites of interest

```{r}
# List MFs that correspond to putative peptide IDs

peptide.mf <- c("C10H16N2O4", "C10H18N2O5", "C10H19N3O4", "C10H20N2O3", "C11H20N2O3", "C11H20N2O5", "C11H22N2O3", "C11H23N5O3", "C12H16N2O3", "C12H25N3O3", "C13H18N2O4", "C13H23N3O4", "C14H20N2O3", "C14H20N2O4", "C15H22N2O3", "C15H22N2O4", "C16H21N3O3", "C8H14N2O4", "C8H16N2O3", "C9H17N3O4", "C9H17N3O4", "C9H18N2O3", "C9H18N2O4", "C17H31N3O4", "C16H29N3O4")

# Remove peptides

index <- which(names(mf.subset) %in% peptide.mf)  
  
no.pep.subset.mf <- mf.subset[-index]

# List InChiKeys corresponding to MFs of putative peptides
index.inchi <- c()

for(inchi in 1:length(inchi.subset)){
  
  if(any(inchi.subset[[inchi]]$MF %in% peptide.mf)){
    
    index.inchi <- c(index.inchi, inchi)
    
  }
}

# Remove peptides
  
no.pep.subset.inchi <- inchi.subset[-index.inchi]

```

# Heatmap of figures significant in at least 1 analysis in all 8 labs

Plot all features in all labs per MF (1 plot per MF) as a function of sample group

```{r}
# heatmap of each MF

subset.heatmap <- function(mf.data){
  # mf.data = data frame of melted, scaled abundance values, sorted by MF or InChi for each lab
  
  # sort data by lab
  
  mf.data$lab <- factor(mf.data$lab, levels = c("Lab8", "Lab7", "Lab6", "Lab5", "Lab4", "Lab3", "Lab2", "Lab1"))
  mf.data <- mf.data[order(mf.data$lab),]
  
  # set order for groups
  
  mf.data$sample.group <- factor(mf.data$sample.group, levels = c("a", "b", "c", "ab", "ac", "bc", "abc"))
  
  # include whether feature was retained in the feature name
  
  mf.data$feature <- paste0(mf.data$retained, "-", mf.data$feature)
  
  # plot
    ## abundance data
      
    abundance.heat <- ggplot(mf.data, aes(sample.group, feature, fill = value)) +
                          geom_tile() +
                          labs(title = paste0(mf.data$MF, " (", gsub(" .*", "", mf.data$Name), ")"), 
                               x = "sample group",
                               y = "") +
                          scale_fill_gradient(low = "white", high = "darkblue") +
                          theme_classic() +
                          theme(axis.text.y = element_text(size = 5)) +
                          scale_y_discrete(labels = function(x) gsub("-.*$", "", x))
    
    ## annotation data
    
    annot.data <- mf.data %>% distinct(feature, lab)
    
    lab.heat <- ggplot(annot.data, aes(x = 1, y = feature, fill = lab)) +
                    geom_tile() +
                    labs(x = "", y = "features by lab") +
                    scale_fill_brewer(type = "qual", palette = 2) +
                    theme_classic() +
                    theme(axis.text.y = element_blank(), axis.text.x = element_blank(),
                          axis.ticks = element_blank(),
                          legend.position = "none",
                          legend.key.size = unit(2, "mm"),
                          line = element_blank())
    
    ## combine plots
    
    heat <- (lab.heat | abundance.heat) + plot_layout(ncol = 2, nrow = 1, widths = unit(c(2, 5), c("mm", "null")))
    
  
  return(heat)
  }

# heatmap

## All overlap MF (26 MFs)

    heatmaps.all <- lapply(mf.subset, function(x) subset.heatmap(x))
    
    heat.arrange.all <- ggarrange(plotlist = heatmaps.all, nrow = 5, ncol = 8, common.legend = F)
    
    ggsave("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/biomarker_heatmap_all.png", heat.arrange.all, width = 30, height = 10, dpi = 300, units = "in", limitsize = F)
 
## No peptides (11 MFs)
    
    heatmaps.nopep <- lapply(no.pep.subset.mf, function(x) subset.heatmap(x))
    
    heat.arrange.nopep <- ggarrange(plotlist = heatmaps.nopep, nrow = 5, ncol = 3, common.legend = F)
    
    ggsave("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/biomarker_heatmap_noPeptides.png", heat.arrange.nopep, width = 16, height = 16, dpi = 300, units = "in", limitsize = F)   
    
    
## InChiKey (22 InChis)
    
    heatmaps.inchi <- lapply(inchi.subset, function(x) subset.heatmap(x))
    
    heat.arrange.inchi <- ggarrange(plotlist = heatmaps.inchi, nrow = 6, ncol = 6, common.legend = F)
    
    ggsave("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/biomarker_heatmap_inchi.png", heat.arrange.inchi, width = 30, height = 11, dpi = 300, units = "in", limitsize = F)   
    
## InChiKey no peptides (9 MFs)
    
    heatmaps.inchi.nopep <- lapply(no.pep.subset.inchi, function(x) subset.heatmap(x))
    
    heat.arrange.inchi.nopep <- ggarrange(plotlist = heatmaps.inchi.nopep, nrow = 3, ncol = 4, common.legend = F)
    
    ggsave("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/biomarker_heatmap_inchi_noPeptides.png", heat.arrange.inchi.nopep, width = 20, height = 15, dpi = 300, units = "in", limitsize = F)   

    
# legend of lab colors
    lab.heat.data <- no.pep.subset.inchi$KDXKERNSBIXSRK %>% distinct(feature, lab)
      
    lab.heat <- ggplot(lab.heat.data, aes(x = 1, y = feature, fill = lab)) +
                    geom_tile() +
                    labs(x = "", y = "features by lab") +
                    scale_fill_brewer(type = "qual", palette = 2) +
                    theme_classic() +
                    theme(axis.text.y = element_blank(), axis.text.x = element_blank(),
                          axis.ticks = element_blank(),
                          legend.key.size = unit(2, "mm"),
                          line = element_blank())
    
    ggsave("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/biomarker_heatmap_labLegend.png", lab.heat, dpi = 300)  
```

# Upset of MFs that met cutoffs

```{r}
# Prep upset data

## only use results that meet cutoffs in all 3 analyses

top.results <- results.df[which(results.df$total.sig == 3),]

## create a list of each lab's MFs

### function to extract MFs for each lab
lab.mf <- function(results.df, lab){
  
  result <- unique(results.df$MF[which(results.df$lab == lab)])
  
  return(result)
  
}

### apply to each lab
labs <- unique(top.results$lab)
mf.upset.data <- lapply(labs, function(x) lab.mf(results.df, x))
names(mf.upset.data) <- labs

# Plot upset

png(filename = "/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/biomarker_upset.png", width = 32, height = 16, units = "in", res = 300)
upset(fromList(mf.upset.data), nsets = 8, nintersects = NA, sets = c("Lab1", "Lab2", "Lab3", "Lab4", "Lab5", "Lab6", "Lab7", "Lab8"), keep.order = T, text.scale = 3)

```

# Boxplot of significant features in at least 1 analysis in all 8 labs

Plot individual features as facets of plot: 1 plot per MF per lab

```{r}
facet.boxplot <- function(mf.data){
  # mf.data = data frame of melted, scaled abundance values, sorted by MF for each lab

    # set group colors
      group.colors <- c("red","gold","blue","orange","purple","darkgreen","brown")
      names(group.colors) <- c("a", "b", "c", "ab", "ac", "bc", "abc")
  
    # adjust sample group order
      mf.data$sample.group <- factor(mf.data$sample.group, levels = c("a", "b", "c", "ab", "ac", "bc", "abc"))
      
  ggplot(mf.data, aes(x = sample.group, y = value, fill = sample.group)) +
    geom_boxplot() +
    theme_classic() +
    scale_fill_manual(values = group.colors) +
    facet_wrap(feature~lab) +
    labs(title = unique(mf.data$MF))
  
}

box.list <- lapply(mf.subset, function(x) facet.boxplot(x))

pdf("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/boxplots_Mffacet.pdf", width = 8.5, height = 11)

for (i in 1:length(box.list)) {
  print(box.list[[i]])
}

dev.off()

#ggsave("/Users/chatelaineha/Documents/Beecher_Collaboration/Writing/data_code/outputs/biomarker_heatmap.png", heat.arrange, width = 20, height = 10, dpi = 300, units = "in")

```

